---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Loading Data ... 
**You can also directly download data from this [Link] [1] ** 
```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r dataload,echo = TRUE}
# Data Science Specialization: Reproducible Research
# Course Project 1
# Vidit Agarwal
################################################################################

rm(list = ls())
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(lattice)

zipFile <- "Data_PeerAssign1.zip"
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

#Setting up folder & files
if(!file.exists("activity.csv")) {
  download.file(fileUrl, destfile=zipFile, method="curl")
  unzip(zipFile)
  file.remove(zipFile)
}

ActivityData <- read.csv("activity.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) 
ActivityData$date <- as.Date(ActivityData$date, "%Y-%m-%d")
```

 

**Lets observe the structure of file**
```{r str}
 str(ActivityData)  
```
 
```{r head, echo = TRUE}
 head(ActivityData)  
```


## What is mean total number of steps taken per day?

*Filtering out missing values in dataset and plotting Histogram*  
```{r totalstep} 
Totalsteps_Day <- ActivityData %>% filter(!is.na(steps)) %>% 
                    group_by(date) %>% summarise( total_steps = sum(steps))
ggplot(Totalsteps_Day, aes(x = total_steps)) + 
               geom_histogram(fill = "grey" ,color = "black", binwidth = 1000) + 
               labs(title="Histogram of Steps Taken per Day", 
                     x = "Number of Steps/Day", y = "Number of days") +
               theme_bw()
```

```{r}
steps_mean   <- as.integer(mean(Totalsteps_Day$total_steps, na.rm=TRUE))
steps_median <- median(Totalsteps_Day$total_steps, na.rm=TRUE)
```
 Therefore, average steps taken per day = **`r steps_mean`**, and median steps = **`r steps_median`**  


## What is the average daily activity pattern?

*Filtering out missing values in dataset and plotting time series plot*  
```{r totalstep2}
  Avgsteps_Interval <- ActivityData %>% filter(!is.na(steps)) %>% 
                       group_by(interval) %>% summarise( avg_steps = mean(steps))
  g <- ggplot(Avgsteps_Interval, aes(x = interval , y = avg_steps)) 
  g + geom_line() + 
      labs(title="'Time Series plot'/n Average Steps per interval", 
          x = "5-Minute Interval", y = "Number of steps") +
      theme_bw()
```

```{r}
best_interval <- as.integer(Avgsteps_Interval[which.max(Avgsteps_Interval$avg_steps),][1])
maxsteps <- as.integer(Avgsteps_Interval[which.max(Avgsteps_Interval$avg_steps),][2])
```
*Time interval in which steps were taken = **`r best_interval`**, steps taken = **`r maxsteps` ** * 

## Imputing missing values

* **Note**: Values will be imputed using mean of respective intervals* 

```{r impute}
  Total_NAs <- sum(is.na(ActivityData$steps))
  Imputed <- merge(ActivityData, Avgsteps_Interval, by = "interval" , by.all = F) 
  Imputed$steps[is.na(Imputed$steps)] <- Imputed$avg_steps[is.na(Imputed$steps)]
  Imputed <- Imputed[names(ActivityData)]

  Totalsteps2_Day <- Imputed%>% filter(!is.na(steps)) %>% 
                     group_by(date) %>% 
                     summarise( total_steps = sum(steps))
  
  ggplot(Totalsteps2_Day, aes(x = total_steps)) + 
  geom_histogram(fill = "grey" ,color = "black", binwidth = 1000) + 
  labs(title="Histogram of Steps Taken per Day", 
       x = "Number of Steps/Day", y = "Number of days") +
  theme_bw()
```

```{r}
steps_mean   <- as.integer(mean(Totalsteps2_Day$total_steps, na.rm=TRUE))
steps_median <- as.integer(median(Totalsteps2_Day$total_steps, na.rm=TRUE))
```
 Therefore, average steps taken per day = **`r steps_mean`**, and median steps = **`r steps_median`**  
 


## Are there differences in activity patterns between weekdays and weekends?

```{r impute2}
Imputed$weekday <- as.factor(ifelse(weekdays(Imputed$date) %in% 
                  c("Sunday", "Saturday"),"Weekend","Weekday"))
Avgsteps2_Interval <- Imputed  %>% group_by(weekday, interval) %>% summarise( avg_steps = mean(steps))

with( Avgsteps2_Interval, xyplot( avg_steps ~ interval | weekday, type = "l",layout = c(1,2), 
main = "'Time Series Plot'\n 5-Minute Interval & Average Steps Taken\n Averaged across weekday or weekend days",
xlab = "5-Min Interval", ylab = "Average Number of Steps"))
```



[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  "Link"

